# Time series / case-crossover study designs 

We'll start by exploring common characteristics in time series data for 
environmental epidemiology. In the first have of the class, we're 
focusing on a very specific type of study---one that leverages large-scale
vital statistics data, collected at a regular time scale (e.g., daily), 
combined with large-scale measurements of a climate-related exposure, with 
the goal of estimating the typical relationship between the level of the 
exposure and risk of a health outcome. For example, we may have daily 
measurements of particulate matter pollution for a city, measured daily at a set
of Environmental Protection Agency (EPA) monitors. We want to investigate
how risk of cardiovascular mortality changes in the city from day to day 
in association with these pollution levels. If we have daily counts of the
number of cardiovascular deaths in the city, we can create a statistical 
model that fits the exposure-response association between particulate matter
concentration and daily risk of cardiovascular mortality. These statistical 
models---and the type of data used to fit them---will be the focus of the 
first part of this course.

## Readings

The required readings for this chapter are: 

- @bhaskaran2013time This paper provides an overview of time series regression
in environmental epidemiology.
- @vicedo2019hands This paper provides a tutorial of all the steps for a 
projecting of health impacts of temperature extremes under climate change. 
One of the steps is to fit the exposure-response association using present-day data
(the section on "Estimation of Exposure-Response Associations" in the paper). 
In this chapter, we will go into details on that step, and that section of the paper
is the only required reading for this chapter. Later in the class, we'll
look at other steps covered in this paper. Supplemental material for this paper is 
available to download by 
clicking http://links.lww.com/EDE/B504. You will need the data in this supplement
for the exercises for class.

The following are supplemental readings (i.e., not required, but may be of
interest) associated with the material in this chapter: 

- @armstrong2012commentary Commentary that provides context on how epidemiological
research on temperature and health can help inform climate change policy.
- @armstrong2014conditional This paper describes different data structures for
case-crossover data, as well as how conditional Poisson regression can be used 
in some cases to fit a statistical model to these data.
Supplemental material for this paper is available at
https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/1471-2288-14-122#Sec13.
- @imai2015time Typically, the time series study design covered in this 
chapter is used to study non-communicable health outcomes. This paper discusses
opportunities and limitations in applying a similar framework for infectious
disease.
- @lu2007equivalence Heavier on statistics. This paper shows how, under
conditions often common for environmental epidemiology studies, case-crossover
and time series methods are equivalent.
- @gasparrini2014modeling Heavier on statistics. This provides the statistical 
framework for the distributed lag model for environmental epidemiology time 
series studies.

## Time series data 

Let's start by exploring the type of dataset that can be used for these 
time series--style studies in environmental epidemiology. In the examples
in this chapter, we'll be using data that comes as part of the Supplemental
Material in one of this chapter's required readings, [@vicedo2019hands].
Follow the link for the supplement for this article and then look for the
file "lndn_obs.csv". This is the file we'll use as the example data in this
chapter.

These data are saved in a csv format (that is, a plain text file, with commas
used as the delimiter), and so they can be read into R using the `read_csv`
function from the `readr` package (part of the tidyverse). For example, you can
use the following code to read in these data, assuming you have saved them in a
"data" subdirectory of your current working directory:

```{r message = FALSE}
library(tidyverse) # Loads all the tidyverse packages, including readr
obs <- read_csv("data/lndn_obs.csv")
obs
```

This example dataset shows many characteristics that are common for datasets for
time series studies in environmental epidemiology. Time series data are essentially 
a sequence of data points repeatedly taken over a certain time interval (e.g., day,
week, month etc). General characteristics of time series data for environmental epidemiology studies are:

- Observations are given at an aggregated level. For example, instead of 
individual observations for each person in London, the `obs` data give 
counts of deaths throughout London. The level of aggregation is often determined 
by geopolitical boundaries, for example, counties or ZIP codes in the US.
- Observations are given at regularly spaced time steps over a period. In the
`obs` dataset, the time interval is day. Typically, values will be provided 
continuously over that time period, with observations for each time interval. 
Occasionally, however, the time series data may only be available for 
particular seasons (e.g., only warm season dates for an ozone study), or
there may be some missing data on either the exposure or health outcome over
the course of the study period.
- Observations are available at the same time step (e.g., daily) for (1) the health outcome, (2) the environmental
exposure of interest, and (3) potential time-varying confounders. In the `obs`
dataset, the health outcome is mortality (from all causes; sometimes, the health
outcome will focus on a specific cause of mortality or other health outcomes such 
as hospitalizations or emergency room visits). Counts are given for everyone in 
the city for each day (`all` column), as well as for specific age categories 
(`all_0_64` for all deaths among those up to 64 years old, and so on). The 
exposure of interest in the `obs` dataset is temperature, and three metrics of 
this are included (`tmean`, `tmin`, and `tmax`). Day of the week is one 
time-varying factor that could be a confounder, or at least help explain 
variation in the outcome (mortality). This is included through the `dow` variable 
in the `obs` data. Sometimes, you will also see a marker for holidays included 
as a potential time-varying confounder, or other exposure variables (temperature 
is a potential confounder, for example, when  investigating the relationship 
between air pollution and mortality risk). 
- Multiple metrics of an exposure and / or multiple health outcome counts 
may be included for each time step. In the `obs` example, three metrics of 
temperature are included (minimum daily temperature, maximum daily temperature, 
and mean daily temperature). Several counts of mortality are included, providing
information for specific age categories in the population. The different metrics 
of exposure will typically be fit in separate models, either as a sensitivity 
analysis or to explore how exposure measurement affects epidemiological results.
If different health outcome counts are available, these can be modeled in separate
statistical models to determine an exposure-response function for each outcome.

## Exploratory data analysis

When working with time series data, it is helpful to start with some exploratory
data analysis. This type of time series data will often be secondary data---it is 
data that was previously collected, as you are re-using it. Exploratory data analysis
is particularly important with secondary data like this. For primary data that you 
collected yourself, following protocols that you designed yourself, you will often
be very familiar with the structure of the data and any quirks in it by the time
you are ready to fit a statistical model. With secondary data, however, you will 
typically start with much less familiarity about the data, how it was collected, and
any potential issues with it, like missing data and outliers. 

Exploratory data analysis can help you become familiar with your data. You can use
summaries and plots to explore the parameters of the data, and also to identify 
trends and patterns that may be useful in designing an appropriate statistical 
model. For example, you can explore how values of the health outcome are distributed, 
which can help you determine what type of regression model would be appropriate, and
to see if there are potential confounders that have regular relationships with 
both the health outcome and the exposure of interest. You can see how many 
observations have missing data for the outcome, the exposure, or confounders of 
interest, and you can see if there are any measurements that look unusual. 
This can help in identifying quirks in how the data were recorded---for example, 
in some cases ground-based weather monitors use -99 or -999 to represent missing
values, definitely something you want to catch and clean-up in your data 
(replacing with R's `NA` for missing values) before fitting a statistical model!

The following applied exercise will take you through some of the
questions you might want to answer through this type of exploratory analysis. 
In
general, the `tidyverse` suite of R packages has loads of tools for exploring and
visualizing data in R.
The `lubridate` package from the `tidyverse`, for example, is an excellent tool for working with date-time data
in R, and time series data will typically have at least one column with the 
timestamp of the observation (e.g., the date for daily data).
You may find it worthwhile to explore this package some more. There is a helpful
chapter in @wickham2016r, https://r4ds.had.co.nz/dates-and-times.html, as well
as a cheatsheet at
https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_lubridate.pdf. For 
visualizations, if you are still learning techniques in R, two books
you may find useful are
@healy2018data (available online at https://socviz.co/) and @chang2018r
(available online at http://www.cookbook-r.com/Graphs/).

*Applied: Exploring time series data*

Read the example time series data into R and explore it to answer the following
questions: 

1. What is the study period for the example `obs` dataset? (i.e., what 
dates / years are covered by the time series data?)
2. Are there any missing dates (i.e., dates with nothing recorded) within this time
period? Are there any recorded dates where health outcome measurements are missing?
Any where exposure measurements are missing?
3. Are there seasonal trends in the exposure? In the outcome?
4. Are there long-term trends in the exposure? In the outcome?
5. Is the outcome associated with day of week? Is the exposure associated
with day of week? 

Based on your exploratory analysis in this section, talk about the potential
for confounding when these data are analyzed to estimate the association between
daily temperature and city-wide mortality. Is confounding by seasonal trends
a concern? How about confounding by long-term trends in exposure and mortality?
How about confounding by day of week?

*Applied exercise: Example code*

1. **What is the study period for the example `obs` dataset? (i.e., what 
dates / years are covered by the time series data?)**

In the `obs` dataset, the date of each observation is included in a column called
`date`. The data type of this column is "Date"---you can check this by using 
the `class` function from base R:

```{r}
class(obs$date)
```

Since this column has a "Date" data type, you can run some mathematical function
calls on it. For example, you can use the `min` function from base R to get the
earliest date in the dataset and the `max` function to get the latest. 

```{r}
min(obs$date)
max(obs$date)
```

You can also run the `range` function to get both the earliest and latest dates
with a single call:

```{r}
range(obs$date)
```

This provides the range of the study period for these data. One interesting point
is that it's not a round set of years---instead, the data ends during the summer
of the last study year. This doesn't present a big problem, but is certainly 
something to keep in mind if you're trying to calculate yearly averages of 
any values for the dataset. If you're getting the average of something that varies
by season (e.g., temperature), it could be slightly weighted by the months that
are included versus excluded in the partial final year of the dataset. Similarly, 
if you group by year and then count totals by year, the number will be smaller
for the last year, since only part of the year's included. For example, if you 
wanted to count the total deaths in each year of the study period, it will look 
like they go down a lot the last year, when really it's only because only about half
of the last year is included in the study period: 

```{r}
obs %>% 
  group_by(year) %>% 
  summarize(deaths_in_year = sum(all)) %>% 
  ggplot(aes(x = year, y = deaths_in_year)) +
  geom_line() +
  geom_point() 
```


2. **Are there any missing dates within this time period? Are there any recorded 
dates where health outcome measurements are missing?
Any where exposure measurements are missing?**

There are a few things you should check to answer this question. First
(and easiest), you can check to see if there are any `NA` values within
any of the observations in the dataset. This helps answer the second and third 
parts of the question. The `summary` function will provide
a summary of the values in each column of the dataset, including the count 
of missing values (`NA`s) if there are any: 

```{r}
summary(obs)
```

Based on this analysis, all observations are complete for all dates included
in the dataset. There are no listings for `NA`s for any of the columns, and 
this indicates no missing values in the dates for which there's a row in the data.

However, this does not guarantee that every date between the start date and 
end date of the study period are included in the recorded data. Sometimes, 
some dates might not get recorded at all in the dataset, and the `summary` 
function won't help you determine when this is the case. One common example in 
environmental epidemiology is with ozone pollution data. These are sometimes
only measured in the warm season, and so may be shared in a dataset with all
dates outside of the warm season excluded.

There are a few alternative explorations you can do to check this. Perhaps the
easiest is to check 
the number of days between the start and end date of the study period, and 
then see if the number of observations in the dataset is the same: 

```{r}
# Calculate number of days in study period
obs %>%            # Using piping (%>%) throughout to keep code clear
  pull(date) %>%   # Extract the `date` column as a vector
  range() %>%      # Take the range of dates (earliest and latest)
  diff()           # Calculate time difference from start to finish of study 

# Get number of observations in dataset---should be 1 more than time difference
obs %>% 
  nrow()
```

This indicates that there is an observation for every date over the study period, 
since the number of observations should be one more than the time difference.
In the next question, we'll be plotting observations by time, and typically this 
will also help you see if there are large chunks of missing dates in the data.

3. **Are there seasonal trends in the exposure? In the outcome?**

You can use a simple plot to visualize patterns over time in both the exposure
and the outcome. For example, the following code plots a dot for each daily
temperature observation over the study period. The points are set to a smaller
size (`size = 0.5`) and plotted with some transparency (`alpha = 0.5`) since
there are so many observations.

```{r}
ggplot(obs, aes(x = date, y = tmean)) + 
  geom_point(alpha = 0.5, size = 0.5)
```

There is (unsurprisingly) clear evidence here of a strong seasonal trend in mean temperature, 
with values typically lowest in the winter and highest in the summer. 

You can plot the outcome variable in the same way: 

```{r}
ggplot(obs, aes(x = date, y = all)) + 
  geom_point(alpha = 0.5, size = 0.5)
```

Again, there are seasonal trends, although in this case they are inversed. 
Mortality tends to be highest in the winter and lowest in the summer. Further, the
seasonal pattern is not equally strong in all years---some years it has a much
higher winter peak, probably in conjunction with severe influenza seasons.

Another way to look for seasonal trends is with a heatmap-style visualization, 
with day of year along the x-axis and year along the y-axis. This allows you 
to see patterns that repeat around the same time of the year each year (and 
also unusual deviations from normal seasonal patterns). 

For example, here's a plot showing temperature in each year, where the
observations are aligned on the x-axis by time in year. We're using the `doy`---which stands for "day of year" (i.e., Jan 1 = 1; Jan 2 = 2; ... Dec 31 = 365 as long
as it's not a leap year) as the measure of time in the year. We've reversed
the y-axis so that the earliest years in the study period start at the top
of the visual, then later study years come later---this is a personal style, 
and it would be no problem to leave the y-axis as-is. We've used the 
`viridis` color scale for the fill, since that has a number of features
that make it preferable to the default R color scale, including that it 
is perceptible for most types of color blindness and be printed out in grayscale
and still be correctly interpreted.

```{r message = FALSE, warning = FALSE}
library(viridis)
ggplot(obs, aes(x = doy, y = year, fill = tmean)) + 
  geom_tile() +
  scale_y_reverse() + 
  scale_fill_viridis()
```

From this visualization, you can see that temperatures tend to be higher in the
summer months and lower in the winter months. "Spells" of extreme heat or cold
are visible---where extreme temperatures tend to persist over a period, rather
than randomly fluctuating within a season. You can also see unusual events, like
the extreme heat wave in the summer of 2003, indicated with the brightest
yellow in the plot.

We created the same style of plot for the health outcome. In this case, we
focused on mortality among the oldest age group, as temperature sensitivity
tends to increase with age, so this might be where the strongest patterns are
evident. 

```{r}
ggplot(obs, aes(x = doy, y = year, fill = all_85plus)) + 
  geom_tile() +
  scale_y_reverse() + 
  scale_fill_viridis()
```

For mortality, there tends to be an increase in the winter compared to the summer.
Some winters have stretches with particularly high mortality---these are likely
a result of seasons with strong influenza outbreaks. You can also see on this 
plot the impact of the 2003 heat wave on mortality among this oldest age group---an
unusual spot of light green in the summer.

4. **Are there long-term trends in the exposure? In the outcome?**

Some of the plots we created in the last section help in exploring this 
question. For example, the following plot shows a clear pattern of decreasing
daily mortality counts, on average, over the course of the study period: 

```{r}
ggplot(obs, aes(x = date, y = all)) + 
  geom_point(alpha = 0.5, size = 0.5)
```

It can be helpful to add a smooth line to help detect these longer-term 
patterns, which you can do with `geom_smooth`: 

```{r message = FALSE}
ggplot(obs, aes(x = date, y = all)) + 
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth()
```

You could also take the median mortality count across each year in the 
study period, although you should take out any years without a full year's 
worth of data before you do this, since there are seasonal trends in the
outcome: 

```{r warning = FALSE}
obs %>% 
  group_by(year) %>% 
  filter(year != 2012) %>% # Take out the last year
  summarize(median_mort = median(all)) %>% 
  ggplot(aes(x = year, y = median_mort)) +
  geom_line()
```

Again, we see a clear pattern of decreasing mortality rates in this city over time. 
This means we need to think carefully about long-term time patterns as a potential 
confounder. It will be particularly important to think about this if the exposure
also has a strong pattern over time. For example, air pollution regulations have
meant that, in many cities, there may be long-term decreases in pollution 
concentrations over a study period. 

5. **Is the outcome associated with day of week? Is the exposure associated
with day of week?**

The data already has day of week as a column in the data (`dow`). However, 
this is in a character data type, so it doesn't have the order of weekdays
encoded (e.g., Monday comes before Tuesday). This makes it hard to look for 
patterns related to things like weekend / weekday. 

```{r}
class(obs$dow)
```

We could convert this to a factor and encode the weekday order when we do 
it, but it's even easier to just recreate the column from the `date` column. 
We used the `wday` function from the `lubridate` package to do this---it extracts
weekday as a factor, with the order of weekdays encoded (using a special 
"ordered" factor type):

```{r warning = FALSE, message = FALSE}
library(lubridate)
obs <- obs %>% 
  mutate(dow = wday(date, label = TRUE))

class(obs$dow)
levels(obs$dow)
```

We looked at the mean, median, and 25th and 75th quantiles of the mortality 
counts by day of week: 

```{r warning = FALSE, message = FALSE}
obs %>% 
  group_by(dow) %>% 
  summarize(mean(all), 
            median(all), 
            quantile(all, 0.25), 
            quantile(all, 0.75))
```

Mortality tends to be a bit higher on weekdays than weekends, but it's not 
a dramatic difference. 

We did the same check for temperature:

```{r message = FALSE, warning = FALSE}
obs %>% 
  group_by(dow) %>% 
  summarize(mean(tmean), 
            median(tmean), 
            quantile(tmean, 0.25), 
            quantile(tmean, 0.75))
```

In this case, there does not seem to be much of a pattern by weekday. 

You can also visualize the association using boxplots:

```{r}
ggplot(obs, aes(x = wday(date, label = TRUE), y = all)) + 
  geom_boxplot()
```

You can also try violin plots---these show the full distribution better than
boxplots, which only show quantiles. 

```{r}
ggplot(obs, aes(x = dow, y = all)) + 
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75))
```

All these reinforce that there are some small differences in weekend versus weekday
patterns for mortality. There isn't much pattern by weekday with temperature, so 
in this case weekday is unlikely to be a confounder (the same is not true with 
air pollution, which often varies based on commuting patterns and so can have 
stronger weekend/weekday differences). However, since it does help some in explaining
variation in the health outcome, it might be worth including in our models anyway, 
to help reduce random noise.

## Fitting models

One of the readings for this week, @vicedo2019hands, includes a section 
on fitting exposure-response functions to describe the association between 
daily mean temperature and mortality risk. This article includes example
code in its supplemental material, with code for fitting the model to 
these time series data in the file named "01EstimationERassociation.r".
The model may at first seem complex, but it is made up of a number of 
fairly straightforward pieces (although some may initially seem complex): 

- The model framework is a *generalized linear model (GLM)*
- This GLM is fit assuming an *error distribution* and a *link function*
appropriate for count data 
- The GLM is fit assuming an *error distribution* that is also appropriate for
data that may be *overdispersed*
- The model includes control for day of the week by including a *categorical
variable*
- The model includes control for long-term and seasonal trends by including
a *spline* (in this case, a *natural cubic spline*) for the day in the study
- The model fits a flexible, non-linear association between temperature 
and mortality risk also using a spline
- The model fits a flexible non-linear association between temperature on 
a series of preceeding days and current day and mortality risk on the 
current day using a *distributed lag approach*
- The model jointly describes both of the two previous non-linear associations
by fitting these two elements through one construct in the GLM, a 
*cross-basis term*

In this section, we will work through the elements, building up the code to 
get to the full model that is fit in @vicedo2019hands.

*Fitting a GLM to time series data*

The GLM framework unites a number of types of regression models you may have
previously worked with. One basic regression model that can be fit within this
framework is a linear regression model. However, the framework also allows you
to also fit, among others, logistic regression models (useful when the outcome
variable can only take one of two values, e.g., success / failure or alive /
dead), Poisson regression models (useful when the outcome variable is a count or 
rate).

This generalized framework brings some unity to these different types of 
regression models. From a practical standpoint, it has allowed software 
developers to easily provide a common interface to fit these types of models. 
In R, the common function call to fit GLMs is `glm`. 

Within the GLM framework, the elements that separate different regression models
include the link function and the error distribution. The error distribution
encodes the assumption you are enforcing about how the errors after fitting the
model are distributed. If the outcome data are normally distributed (a.k.a.,
follow a Gaussian distribution), after accounting for variance explained in the
outcome by any of the model covariates, then a linear regression model may be
appropriate. For count data---like numbers of deaths a day---this is unlikely, 
unless the average daily mortality count is very high (count data tend to 
come closer to a normal distribution the further their average gets from 
0). For binary data---like whether each person in a study population died on 
a given day or not---normally distributed errors are also unlikely. Instead, 
in these two cases, it is typically more appropriate to fit GLMs with 
Poisson and binomial "families", respectively, where the family designation 
includes an appropriate specification for the variance when fitting the model
based on these outcome types. 

The other element that distinguishes different types of regression within 
the GLM framework is the link function. The link function applies a transformation
on the combination of independent variables in the regression equation 
when fitting the model. With normally distributed data, an *identity link*
is often appropriate---with this link, the combination of independent variables
remain unchanged (i.e., keep their initial "identity"). With count data, a 
*log link* is often more appropriate, while with binomial data, a *logit link*
is often used. 

Finally, data will often not perfectly adhere to assumptions. For example, the
Poisson family of GLMs assumes that variance follows a Poisson distribution 
(The probability mass function for Poisson distribution $X \sim {\sf Poisson}(\mu)$ is denoted by $f(k;\mu)=Pr[X=k]= \displaystyle \frac{\mu^{k}e^{-\mu}}{k!}$, where 
$k$ is the number of occurences, and $\mu$ is equal to the expected number of 
cases). With this distribution, the variance is equal to the mean ($\mu=E(X)=Var(X)$). With real-life data, this assumption is often not valid, and in many cases the variance in  real life count data is larger than the mean. This can be accounted  for when fitting a GLM by setting an error distribution that does not require the variance to equal the mean---instead, both a mean value and something like a 
variance are estimated from the data, assuming an overdispersion parameter $\phi$ 
so that $Var(X)=\phi E(X)$. In environmental epidemiology, time series 
are often fit to allow for this overdispersion. This is because if the data are overdispersed but the model does not account for this, the standard errors on the 
estimates of the model parameters may be artificially small. If the data are not overdispersed ($\phi=1$), the model will identify this when being fit to the data, 
so it is typically better to prefer to allow for overdispersion in the model 
(if the size of the data were small, you may want to be parsimonious and avoid 
unneeded complexity in the model, but this is typically not the case with time 
series data). 

In the next section, you will work through the steps of developing a GLM to fit
the example dataset `obs`. For now, you will only fit a linear association
between mean daily temperature and mortality risk, eventually including control
for day of week. In later work, especially the next chapter, we will build up
other components of the model, including control for the potential confounders
of long-term and seasonal patterns, as well as advancing the model to fit
non-linear associations, distributed by time, through splines, a distributed lag
approach, and a cross-basis term.

*Applied: Fitting a GLM to time series data*

In R, the function call used to fit GLMs is `glm`. Most of you have likely 
covered GLMs, and ideally this function call, in previous courses. If you are
unfamiliar with its basic use, you will want to refresh yourself on this 
topic. [Add some online resources that go over basics of GLMs in R.]

1. Fit a GLM to estimate the association between mean daily temperature (as the
independent variable) and daily mortality count (as the dependent variable),
first fitting a linear regression. (Since the mortality data are counts, we will
want to shift to a different type of regression within the GLM framework, but
this step allows you to develop a simple `glm` call, and to remember where to
include the data and the independent and dependent variables within this
function call.)
2. Change your function call to fit a regression model in the Poisson family.
3. Change your function call to allow for overdispersion in the outcome data 
(daily mortality count). How does the estimated coefficient for temperature
change between the model fit for #2 and this model? Check both the central 
estimate and its estimated standard error.
4. Change your function call to include control for day of week. 
 
*Applied exercise: Example code*

1. **Fit a GLM to estimate the association between mean daily temperature (as the
independent variable) and daily mortality count (as the dependent variable),
first fitting a linear regression.**

This is the model you are fitting:

$Y_{t}=\beta_{0}+\beta_{1}X1_{t}+\epsilon$ 
              
where $Y_{t}$ is the mortality count on day $t$, $X1_{t}$ is the mean temperature 
for day $t$ and $\epsilon$ is the error term. Since this is a linear model we are 
assuming a Gaussian error distribution $\epsilon \sim {\sf N}(0, \sigma^{2})$,
where $\sigma^{2}$ is the variance not explained by the covariates (here just 
temperature). 

To do this, you will use the `glm` call. If you would like to save model fit
results to use later, you assign the output a name as an R object
(`mod_linear_reg` in the example code). If your study data are in a dataframe, 
you can specify these data in the `glm` call with the `data` parameter. 
Once you do this, you can use column names directly in the model formula. 
In the model formula, the dependent variable is specified first (`all`, the 
column for daily mortality counts for all ages, in this example), followed
by a tilde (`~`), followed by all independent variables (only `tmean` in this
example). If multiple independent variables are included, they are joined using
`+`---we'll see an example when we start adding control for confounders later.

```{r}
mod_linear_reg <- glm(all ~ tmean, data = obs)
```

Once you have fit a model and assigned it to an R object, you can explore it
and use resulting values. First, the print method for a regression model
gives some summary information. This method is automatically called if you 
enter the model object's name at the console: 

```{r}
mod_linear_reg
```

More information is printed if you run the `summary` method on the model 
object: 

```{r}
summary(mod_linear_reg)
```

Make sure you are familiar with the information provided from the model object,
as well as how to interpret values like the coefficient estimates and their
standard errors and p-values. These basic elements should have been covered in
previous coursework (even if a different programming language was used to fit
the model), and so we will not be covering them in great depth here, but instead
focusing on some of the more advanced elements of how regression models are
commonly fit to data from time series and case-crossover study designs in
environmental epidemiology. For a refresher on the basics of fitting 
statistical models in R, you may want to check out Chapters 22 through 24 of
@wickham2016r, a book that is available online.

Finally, there are some newer tools for extracting information from model fit 
objects. The `broom` package extracts different elements from these objects
and returns them in a "tidy" data format, which makes it much easier to use
the output further in analysis with functions from the "tidyverse" suite of
R packages. These tools are very popular and powerful, and so the `broom` tools
can be very useful in working with output from regression modeling in R. 

The `broom` package includes three main functions for extracting data from 
regression model objects. First, the `glance` function returns overall data 
about the model fit, including the AIC and BIC:

```{r}
library(broom)
glance(mod_linear_reg)
```

The `tidy` function returns data at the level of the model coefficients, 
including the estimate for each model parameter, its standard error, test 
statistic, and p-value.

```{r}
tidy(mod_linear_reg)
```

Finally, the `augment` function returns data at the level of the original 
observations, including the fitted value for each observation, the residual
between the fitted and true value, and some measures of influence on the model
fit.

```{r}
augment(mod_linear_reg)
```

One way you can use `augment` is to graph the fitted values for each observation
after fitting the model: 

```{r}
mod_linear_reg %>% 
  augment() %>% 
  ggplot(aes(x = tmean)) + 
  geom_point(aes(y = all), alpha = 0.4, size = 0.5) + 
  geom_line(aes(y = .fitted), color = "red") + 
  labs(x = "Mean daily temperature", y = "Log(Expected mortality count)")
```

For more on the `broom` package, including some excellent examples of how it
can be used to streamline complex regression analyses, see @robinson2014broom. 
There is also a nice example of how it can be used in one of the chapters of
@wickham2016r, available online at https://r4ds.had.co.nz/many-models.html. 

2. **Change your function call to fit a regression model in the Poisson family.**

A linear regression is often not appropriate when fitting a model where the 
outcome variable provides counts, as with the example data. A Poisson regression
is often preferred. 

For a count distribution were $Y \sim {\sf Poisson(\mu)}$ we typically fit a model
such as 

$g(Y)=\beta_{0}+\beta_{1}X1$, where $g()$ represents the link function, in this 
case a log function so that  $log(Y)=\beta_{0}+\beta_{1}X1$. We can also express 
this as $Y=exp(\beta_{0}+\beta_{1}X1)$.

In the `glm` call, you can specify this with the `family`
parameter, for which "poisson" is one choice.

```{r}
mod_pois_reg <- glm(all ~ tmean, data = obs, family = "poisson")
```

One thing to keep in mind with this change is that the model now uses a 
non-identity link between the combination of independent variable(s) and the 
dependent variable. You will need to keep this in mind when you interpret 
the estimates of the regression coefficients. While the coefficient estimate
for `tmean` from the linear regression could be interpreted as the expected 
increase in mortality counts for a one-unit (i.e., one degree Celsius) increase
in temperature, now the estimated coefficient should be interpreted as the
expected increase in the natural log-transform of mortality count for a one-unit
increase in temperature. 

```{r}
summary(mod_pois_reg)
```

You can see this even more clearly if you take a look at the association between
temperature for each observation and the expected mortality count fit by the 
model. First, if you look at the fitted values without transforming, they 
will still be in a state where mortality count is log-transformed. You can 
see by looking at the range of the y-scale that these values are for the log
of expected mortality, rather than expected mortality, and that the fitted
association is linear: 

```{r}
mod_pois_reg %>% 
  augment() %>% 
  ggplot(aes(x = tmean)) + 
  geom_point(aes(y = log(all)), alpha = 0.4, size = 0.5) + 
  geom_line(aes(y = .fitted), color = "red") + 
  labs(x = "Mean daily temperature", y = "Log(Expected mortality count)")
```

You can use exponentiation to transform the fitted values back to just be the
expected mortality count based on the model fit. Once you make this
transformation, you can see how the link in the Poisson family specification
enforced a curved relationship between mean daily temperature and the
untransformed expected mortality count.

```{r}
mod_pois_reg %>% 
  augment() %>% 
  ggplot(aes(x = tmean)) + 
  geom_point(aes(y = all), alpha = 0.4, size = 0.5) + 
  geom_line(aes(y = exp(.fitted)), color = "red") + 
  labs(x = "Mean daily temperature", y = "Expected mortality count")
```

3. **Change your function call to allow for overdispersion in the outcome data 
(daily mortality count). How does the estimated coefficient for temperature
change between the model fit for #2 and this model? Check both the central 
estimate and its estimated standard error.**

In the R `glm` call, there is a family that is similar to Poisson (including
using a log link), but that allows for overdispersion. You can specify it 
with the "quasipoisson" choice for the `family` parameter in the `glm` call:

```{r}
mod_ovdisp_reg <- glm(all ~ tmean, data = obs, family = "quasipoisson")
```

When you use this family, there will be some new information in the summary
for the model object. It will now include a dispersion parameter ($\phi$). If this 
is close to 1, then the data were close to the assumed variance for a Poisson
distribution (i.e., there was little evidence of overdispersion). In the 
example, the overdispersion is around 5, suggesting the data are overdispersed
(this might come down some when we start including independent variables that
explain some of the variation in the outcome variable, like long-term and
seasonal trends). 

```{r}
summary(mod_ovdisp_reg)
```

If you compare the estimates of the temperature coefficient from the Poisson
regression with those when you allow for overdispersion, you'll see something
interesting: 

```{r}
tidy(mod_pois_reg) %>% 
  filter(term == "tmean")
tidy(mod_ovdisp_reg) %>% 
  filter(term == "tmean")
```

The central estimate (`estimate` column) is very similar. However, the estimated
standard error is larger when the model allows for overdispersion. This
indicates that the Poisson model was too simple, and that its inherent 
assumption that data were not overdispersed was problematic. If you naively used
a Poisson regression in this case, then you would estimate a confidence 
interval on the temperature coefficient that would be too narrow. This could 
cause you to conclude that the estimate was statistically significant when 
you should not have (although in this case, the estimate is statistically 
significant under both models). 

4. **Change your function call to include control for day of week.**

Day of week is included in the data as a categorical variable, using a 
data type in R called a factor. You are now essentially fitting this model:

$log(Y)=\beta_{0}+\beta_{1}X1+\gamma^{'}X2$,

where $X2$ is a categorical variable for day of the week and $\gamma^{'}$ 
represents a vector of parameters associated with each category.

It is pretty straightforward to include factors as independent variables in calls 
to `glm`: you just add the column name to the list of other independent variables 
with a `+`. In this case, we need to do one more step: earlier, we added order to
`dow`, so it would "remember" the order of the week days (Monday before Tuesday, 
etc.). However, we need to strip off this order before we include the factor in 
the `glm` call. One way to do this is with the `factor` call, specifying 
`ordered = FALSE`. Here is the full call to fit this model:

```{r}
mod_ctrl_dow <- glm(all ~ tmean + factor(dow, ordered = FALSE), 
                    data = obs, family = "quasipoisson")
```

When you look at the summary for the model object, you can see that the 
model has fit a separate model parameter for six of the seven weekdays. The one
weekday that isn't fit (Sunday in this case) serves as a baseline ---these 
estimates specify how the log of the expected mortality count is expected to 
differ on, for example, Monday versus Sunday (by about 0.03), if the temperature 
is the same for the two days. 

```{r}
summary(mod_ctrl_dow)
```

You can also see from this summary that the coefficients for the day of the
week are all statistically significant. Even though we didn't see a big 
difference in mortality counts by day of week in our exploratory analysis, 
this suggests that it does help explain some variance in mortality observations
and will likely be worth including in the final model. 

The model now includes day of week when fitting an expected mortality count
for each observation. As a result, if you plot fitted values of expected
mortality versus mean daily temperature, you'll see some "hoppiness" in the 
fitted line: 

```{r}
mod_ctrl_dow %>% 
  augment() %>% 
  ggplot(aes(x = tmean)) + 
  geom_point(aes(y = all), alpha = 0.4, size = 0.5) + 
  geom_line(aes(y = exp(.fitted)), color = "red") + 
  labs(x = "Mean daily temperature", y = "Expected mortality count")
```

This is because each fitted value is also incorporating the expected influence
of day of week on the mortality count, and that varies across the observations
(i.e., you could have two days with the same temperature, but different 
expected mortality from the model, because they occur on different days). 

If you plot the model fits separately for each day of the week, you'll see that 
the line is smooth across all observations from the same day of the week:

```{r}
mod_ctrl_dow %>% 
  augment() %>% 
  ggplot(aes(x = tmean)) + 
  geom_point(aes(y = all), alpha = 0.4, size = 0.5) + 
  geom_line(aes(y = exp(.fitted)), color = "red") + 
  labs(x = "Mean daily temperature", y = "Expected mortality count") + 
  facet_wrap(~ obs$dow)
```

*Wrapping up*

At this point, the coefficient estimates suggests that risk of mortality 
tends to decrease as temperature increases. Do you think this is reasonable?
What else might be important to build into the model based on your analysis
up to this point?

## Chapter vocabulary

Each class will start with a vocabulary quiz on a select number of the words
from the chapter's vocabulary list. The vocabulary words for this chapter are: 

- time-series study design
- case-crossover study design
- exposure
- health outcome
- confounder
- study period
- seasonal trends
- long-term trends
- error distribution
- generalized linear model (GLM)
- link function
- overdispersed
- categorical variable
- spline
- natural cubic spline
- distributed lag
- cross-basis term