# Natural experiments

The readings for this chapter are: 

- @bernal2017interrupted (on interrupted time series), with a correction to an equation in the paper at https://academic.oup.com/ije/article/49/4/1414/5900884. Example data and R code for the paper are available [to download](https://oup.silverchair-cdn.com/oup/backfile/Content_public/Journal/ije/49/4/10.1093_ije_dyaa118/1/dyaa118_supplementary_data.zip?Expires=1623897009&Signature=BzYQrBg60cMKHYeDU~OIZYIFuRgEIPwQsWMjzON0dB~fL8y-8x4xdGIJQBBPgDxBIoUIGnjmShVf1jlVqzloo3IldAdVC78TZ~~XseYdJ9c590QRAR6m7mH~VbPe-fCnQSnZF0z2Qw9PZcSGITZeNr4YXPVY-~gtpgBeZiN0MpgEVBLVT5fYhhQBGbp0vxl1bKdUfNtF71fdVJrglkhSG8-M24A07LmAr8jThx4MQmSAzKCxA4VZLRE6To8zC3-rJlxyWiqrSTFsVQM2SN4R6UuxYoRsILRcIAr2sUfqgmaSlxBiYAf71PdGSrnBcXX3l0l7yuAftX5PYTwMKTyxOA__&Key-Pair-Id=APKAIE5G5CRDK6RD3PGA) through a Supplemental Appendix.
- @barone2011effects, the scientific paper highlighted as an example in the tutorial in the previous reading
- @bor2014regression (on interrupted time series)
- @casey2018retirements (on difference-in-differences)
- @mendola2018invited, an Invited Commentary on the previous reading

## Interrupted time series

[Interrupted time series assessing effects of policy/intervention in specific point in time]

```{r message = FALSE, warning = FALSE}
# Load some packages that will likely be useful
library(tidyverse)
library(viridis)
library(lubridate)
library(broom)

# Load and clean the data
obs <- read_csv("data/lndn_obs.csv") %>% 
  mutate(dow = wday(date, label = TRUE)) %>% 
  mutate(time = as.numeric(date) - first(as.numeric(date)))
```

```{r}
london_summer_2012 <- obs %>% 
  filter(ymd("2012-06-01") <= date & date <= ymd("2012-09-30"))

london_olympic_dates <- tibble(date = ymd(c("2012-07-27", "2012-08-12")))

ggplot() + 
  geom_polygon(aes(x = ymd(c("2012-07-27", "2012-08-12", 
                             "2012-08-12", "2012-07-27")), 
                   y = c(Inf, Inf, -Inf, -Inf)), fill = "cyan", alpha = 0.2) + 
  geom_line(data = london_summer_2012, aes(x = date, y = tmean)) + 
  labs(x = "Date", y = "Mean daily temperature")
```

Example data from @bernal2017interrupted: 

```{r message = FALSE}
sicily <- read_csv("data/sicily.csv") %>% 
  mutate(date = paste(year, month, "15"), # Use middle of the month for plotting
         date = ymd(date))
```

Identify dates of the smoking band: 

```{r}
sicily %>% 
  group_by(smokban) %>% 
  slice(c(1, n()))
```


Recreate Figure 1 from @bernal2017interrupted: 

```{r message = FALSE}
ggplot() + 
  geom_polygon(aes(x = ymd(c("2005-01-01", "2006-11-30", 
                         "2006-11-30", "2005-01-01")), 
               y = c(Inf, Inf, -Inf, -Inf)), fill = "lightgray") + 
  geom_point(data = sicily, 
             aes(x = date, y = 10000 * 10 * aces / stdpop), shape = 21) + 
  geom_smooth(data = sicily, 
              aes(x = date, y = 10000 * 10 * aces / stdpop), # Need the extra 10 to line up with Figure in paper---figure out why
              method = "lm", se = FALSE, color = "red", linetype = 3) + 
  labs(x = "Year", y = "Std rate x 10 000") + 
  theme_classic()
```


## Difference-in-differences

[Difference-in differences application for intervention introduced in one point in time]